---
title: "Monkeypox"
output: pdf_document
date: '2022-10-01'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Description of dataset
What is the data source? (1-2 sentences on where the data is coming from, dates included, etc.)
The data is about the Monkey case data is from European Centre for Disease Prevention and Control
The population denominator and census data is from Eurostat

How does the dataset relate to the group problem statement and question?
Is there a difference between monkeypox cases by region in the EU
We will be using the monkeypox case rate, population denominator and region data set to help us find the difference between monkeypox cases by region in the EU.
Is there a difference between country level monkeypox case rates by certain demographic factors?
We will use the monkeypox case, population denominator and census data to see if there is a difference between country level case rates based on certain demographic factors.

\newpage

Import statement
Use appropriate import function and package based on the type of file
Utilize function arguments to control relevant components (i.e. change column types, column names, missing values, etc.)
Document the import process
1, We navigated to the PHW290/phw251_projectdata git repo. We clicked on Code and downloaded the Zip file which we then extracted the csv files from. We imported the relevant datasets into our RStudio files for this project.
```{r}
library(readr)
library(dplyr)
library(tidyverse)
```
```{r}
mp_cases <-read_csv("~/r_for_ph_monkeypox/euro_mpx_cases.csv")
pop_den <- read_csv("~/r_for_ph_monkeypox/euro_pop_denominators.csv")
eu_census <- read_csv("euro_census_stats.csv")
wrld_region <- read_csv("~/r_for_ph_monkeypox/world_country_regions.csv")
```


\newpage
Identify data types for 5+ data elements/columns/variables
Identify 5+ data elements required for your specified scenario. If <5 elements are required to complete the analysis, please choose additional variables of interest in the data set to explore in this milestone.
Utilize functions or resources in RStudio to determine the types of each data element (i.e. character, numeric, factor)
Identify the desired type/format for each variable—will you need to convert any columns to numeric or another type? 
Provide a basic description of the 5+ data elements
Numeric: mean, median, range
Character:  unique values/categories
Or any other descriptives that will be useful to the analysis
```{r}
clean_eu_census <- rename_with(eu_census, ~ tolower((gsub(" ", "_",.x,fixed=TRUE)))) %>%
  select(c(-"flags",-"footnotes"))
```
