---
title: "Monkeypc note"
output: pdf_document
date: '2022-11-06'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
# eu_census_demo_clean <- eu_cens_demo %>%
#   mutate(age_ord = recode(age, "Y_LT15" = "1",
#                               "Y15-29" = "2",
#                               "Y30-49" = "3",
#                               "Y50-64" = "4",
#                               "Y65-84" = "5",
#                               "Y_GE85" = "6"))

# eu_census_demo_clean <- eu_cens_demo %>%
#    mutate(age_cat = case_when(age == "Y_LT15" ~ "Younger than 15",
#                               age == "Y15-29" ~ "Age 15-29",
#                               age == "Y30-49" ~ "Age 30-49",
#                               age == "Y50-64" ~ "Age 50-64",
#                               age == "Y65-84" ~ "Age 65-85",
#                               age == "Y_GE85" ~ "85 and Older"))

# eu_census_demo_clean$age_cat <- factor(eu_census_demo_clean$age_cat,
# levels = c("Younger than 15","Age 15-29","Age 30-49","Age 50-64","Age 65-85","85 and Older"), #"2","3","1","4","5"
# #labels = c("Staten Island","Manhattan",
# #"Bronx", "Queens", "Brooklyn"),
# ordered=TRUE)

```

```{r}
# join_cens_mp <- left_join(eu_cens_demo, join_region_mp, by = c("country_code" = "geo"))%>%
#   group_by(sex, monthly_total_cases)

```

```{r}
 # filter(age_cat == "Younger than 15") %>%
  # mutate(pct_young = case_when (group_pct1 >= 0.50 ~ "Greater than 50% of Population is younger than 15",
  #                              group_pct1 >= 0.30 ~ "Between 30% and 50% is younger than 15",
  #                              group_pct1 < 0.30 ~ "Less than 30% is younger than 15"

#age_mp_cases <- left_join(eu_census_age_demo, join_region_mp, by = c("country_code" = "geo"))# %>%
# group_by(country_code, group_pct, month, `sub-region`)# %>%
#   summarize(cases_by_monthcat = sum(monthly_total_cases, na.rm = T),
#             cat_pop = max(country_pop, na.rm = T)) %>%
#   mutate(size_cat_rate = (cases_by_monthcat / cat_pop) * 1000000) %>%
#   ungroup()

```
eu_census_age_demo <- clean_eu_census %>%
  mutate(age_cat = case_when(age == "Y_LT15" ~ "Younger than 15",
                              age %in% c("Y15-29","Y30-49")  ~ "Between 15 to 49",
                              age %in% c("Y50-64","Y65-84", "Y_GE85") ~ "50 and Older"))%>%
  group_by(country_code, age_cat) %>%
  summarize(group_pop1 = sum(pop, na.rm = T)) %>%
  mutate(country_pop1 = sum(group_pop1, na.rm = T),
         group_pct1 = group_pop1 / country_pop1) %>%
  ungroup() %>%
  filter(age_cat == "Younger than 15") %>%
  mutate(pct_young = case_when (group_pct1 >= 0.50 ~ "Greater than 50% of Population is younger than 15",
                               group_pct1 >= 0.30 ~ "Between 30% and 50% is younger than 15",
                               group_pct1 < 0.30 ~ "Less than 30% is younger than 15"))


  age_mp_cases <- left_join(eu_census_age_demo, join_region_mp, by = c("country_code" = "geo")) %>%
     group_by(country_code, month, `sub-region`,age_cat)%>% 
    summarize(cases_by_monthcat = sum(monthly_total_cases, na.rm = T),
            age_pop = max(group_pop1, na.rm = T)) %>%
  mutate(size_cat_rate = (cases_by_monthcat / age_pop) * 1000000) %>%
  ungroup()

